---
title: "Seurat implementation"
author: "Carlos Ramirez Alvarez"
date: "9/18/2019"
output: html_document
---

## 10X input

The following files are the input of Seurat 10x function.

```{r}
dataPath <- '../data/pbmc3k/filtered_gene_bc_matrices/hg19/'
files <- list.files(dataPath)
files
```

* Inspecting **barcodes.tsv** file.

```{r}
fPath <- paste(dataPath, files[1], sep = '')
read.table(fPath, nrows = 5)
```

barcodes file has 2,700 lines.

```{r}
command <- paste0('wc -l ', fPath)
system(command)
```

* Display **genes.tsv** file content.

```{r}
fPath <- paste(dataPath, files[2], sep = '')
read.table(fPath, nrows = 5)
```

```{r}
command <- paste0('wc -l ', fPath)
system(command)
```

Show **matrix.mtx** file head content.

```{r}
fPath <- paste(dataPath, files[3], sep = '')
substr(readLines(fPath, n = 5),
       start = 1, stop = 50)
```

These files are the output of the [cellRanger](https://support.10xgenomics.com/single-cell-gene-expression/software/pipelines/latest/what-is-cell-ranger) pipeline developed by 10X Genomics. See Feature [Barcode Matrix](https://support.10xgenomics.com/single-cell-gene-expression/software/pipelines/latest/output/matrices) documentations for the description of these files.

## Loading data into Seurat


```{r warning=FALSE, message=FALSE}
library(Seurat)

pbmc <- Read10X(data.dir = '../data/pbmc3k/filtered_gene_bc_matrices/hg19/')
str(pbmc)
```

